---
title: "US Unemployment Analysis"
description: |
   An brief analysis of the US employment rate since 2016 using the data from the Federal Reserve Bank of St.Louis.
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(fpp2)
library(stringi)
library(gridExtra)
library(plotly)
library(gganimate)
library(gifski)
```

## Introduction

This is a project I completed during the summer of 2022, I was initially inspired by an Instagram post published by `thedemocrats`. Along with a visual animation showing a decreasing trend, it claiming that as of July 2022, the "Unemployment Rate is Tied  For The Lowest In 50 Years". I am motivated to then dig deeper to see what happened during these 50 years, as well as comparison between the progress made by the current president Joe Biden and the former president Donald Trump. 

## Credit 

This analysis used the `tidyverse`,`gridExtra`, `plotly`, `fpp2` and `stringi` packages. Every functions used in this analysis belongs to the creator of these packages and as well as the creator of R. 

## Seasonal Plot Making

```{r}
# load data from FRED
usunemp <- read_csv("./data/unempl.csv",show_col_types = FALSE)
```

The data I used is from the [Unemployment Rate]("https://fred.stlouisfed.org/series/UNRATE") data set provided by [FRED](https://fred.stlouisfed.org/).

This is the raw data from the FRED data set.

```{r echo = FALSE}
usunemp
```

Start the pipe with the data set, first use the `stri_sub` function from the `stringi` package to extract the year value of the `Date` column,use `as.double` to turn the vector type `Date` to numerical vector `Double`. Then use the `filter` function from the `tidyverse` package to filter out the desired year (in this case between 2016 and 2021), Lastly, select the `UNRATE` column.

```{r}
us_unemployment_1 <- usunemp |>
  mutate(DATE = as.double(stri_sub(DATE,0,4))) |>
  filter(DATE >= 2016 & DATE < 2021)|>
  select(UNRATE)
  
```

The resulting tibble is shown below, which contains a total of 60 rows.  
\n(5 years * 12 months/years)

```{r echo = FALSE}
us_unemployment_1
```
Use `ts` function to translate the tibble into a time-series object, note the frequency correspond with the pattern of repetition within the data (months), and using the same start year as set during the previous code chunk.

```{r}
us_unemployment_g1 <-ts(us_unemployment_1,frequency=12,start = 2016)

```

```{r echo = FALSE}
us_unemployment_g1 
```

Lastly, use `ggseasonplot` to visualize the dataset, set `polar = TRUE` will give us the cool polar coordinate in the plot. Adjust the titles and texts with `theme`, `xlab`, `ylab`, and `ggtitle`.

```{r}
plot1 <- ggseasonplot(us_unemployment_g1, polar=TRUE)+
theme(text = element_text(family = "STHeiti"))+
  theme(plot.title = element_text(hjust = 0.5))+
  xlab("Month")+
  ylab("Unemployment Rate") +
  ggtitle("US Unemployment Rate")

```

```{r echo = FALSE}
plot1
```

Similarly, we can create the plot from 2020 to 2022 using the same technique.


```{r}
us_unemployment_2 <- usunemp |>
  mutate(DATE = as.double(stri_sub(DATE,0,4))) |>
  filter(DATE >= 2020 & DATE <= 2022)|>
  select(UNRATE)

us_unemployment_g2 <-ts(us_unemployment_2,frequency=12,start = 2020)

plot2 <- ggseasonplot(us_unemployment_g2, polar=TRUE)+
theme(text = element_text(family = "STHeiti"))+
  theme(plot.title = element_text(hjust = 0.5))+
  xlab("Month")+
  ylab("Unemployment Rate") +
  ggtitle("US Unemployment Rate")

```

```{r echo = FALSE}
plot2
```

## Analysis

Use `grid.arrange` from the `gridExtra` package to put the two plots side by side.

```{r}
grid.arrange(plot1, plot2, ncol=2)
```

The two plots represent the unemployment rate in the United States under the 45th President Donald Trump (Left), and the 46th President Joe Biden (Right). The plot at the bottom shows the historical US unemployment rate since January 1948.

The data have shown that the unemployment rate under the first three years of President Trump have gradually decreases, however, starting in January 2020, when Covid first hit the United States, the unemployment rate skyrocketed to an all time high of 14.7% in April 2022. After the peek in April, the unemployment rate decreases, and during the time of President Biden's inauguration, the US unemployment rate was around 6.4%. After President Biden took office, the unemployment rate continues to drop, as of July 2022, the unemployment rate of the US was at it's lowest point since 1960, at around 3.5%. 


```{r echo = FALSE}
hist_plot <- usunemp|>
  ggplot(aes(x = DATE, y = UNRATE, text = paste("</br>Date:",DATE, '</br> Value:',UNRATE)))+
  geom_line(aes(group = 1),color='darkblue')+
  labs(title = "US unemployment rate (Jan 1948 - Jul 2022)", 
       x = " Date (Monthly)", 
       y = "Unemployment Rate (Percentage)")+
  theme_bw()

ggplotly(hist_plot, tooltip = "text")
```

```{r include = FALSE}
anim_plot <- usunemp|>
  filter(DATE >= '1948-01-01') |>
  ggplot(aes(x = DATE, y = UNRATE))+
  geom_point() +
  geom_line(color='darkblue')+
  transition_reveal(DATE) +
  labs(title = "US unemployment rate (Jan 1948 - Jul 2022)", 
       x = " Date (Monthly)", 
       y = "Unemployment Rate (Percentage)")+
  theme_bw()


animate(anim_plot, duration = 10, renderer = gifski_renderer())
anim_save("usunemp.gif")
```

This analysis is also published on Rpubs, click [here](https://rpubs.com/YuhanWu/us-unemploy) to take a look!

